---
title: "<b>San Francisco Bay Area’s Bike Share System<b>"
date: "4/5/2020"
output: html_document
---

*** 

Source: [datascienceplus.com](https://datascienceplus.com/exploring-san-francisco-bay-areas-bike-share-system/)

*** 

## **Highlights**
* The number of rides per month more than doubled from January to June of 2018.
* Monthly and yearly subscribers have completed more than five times the number of rides compared to customers who hold single-use or day-ride passes.
* The ratio of male to female users is 3:1.
* The median age of users is 35 with a standard deviation of 10.5.
* More than 90% of the rides are under 30 minutes.
* The peak hours across all rides in 2018 are 8 AM and 5 PM, which align with normal working hours.
* The bike share system in San Francisco is popular among those who commute from outside of the city for work.

*** 

## **Libraries**
```{r libraries}
library(dplyr)   # for data manipulation
library(stringr) # for regular expression
library(ggplot2) # for plotting functions
```

***

## **Data Acquisition**

```{r data-acquisition}
# LIst of data file names
file_names = list.files(pattern = ".csv")
file_names

# Read data (this will take a second)
df_list = lapply(file_names,
                 function(x) read.csv(x, stringsAsFactors = FALSE))

# Assign names to data frames in the list
names(df_list) = file_names
```

*** 

## **Data Organization and Cleaning**
### Aggregate Data into a Single Data Frame

```{r aggr-data-into-single-df}
# Extract column names from the first data frame
names(df_list[[1]])

# Cast certain columns as numeric to enable row binding
df_list[[6]]$start_station_id = as.numeric(df_list[[6]]$start_station_id)
df_list[[6]]$end_station_id = as.numeric(df_list[[6]]$end_station_id)
df_list[[7]]$start_station_id = as.numeric(df_list[[7]]$start_station_id)
df_list[[7]]$end_station_id = as.numeric(df_list[[7]]$end_station_id)
df_list[[8]]$start_station_id = as.numeric(df_list[[8]]$start_station_id)
df_list[[8]]$end_station_id = as.numeric(df_list[[8]]$end_station_id)
df_list[[9]]$start_station_id = as.numeric(df_list[[9]]$start_station_id)
df_list[[9]]$end_station_id = as.numeric(df_list[[9]]$end_station_id)
df_list[[10]]$start_station_id = as.numeric(df_list[[10]]$start_station_id)
df_list[[10]]$end_station_id = as.numeric(df_list[[10]]$end_station_id)
df_list[[11]]$start_station_id = as.numeric(df_list[[11]]$start_station_id)
df_list[[11]]$end_station_id = as.numeric(df_list[[11]]$end_station_id)
df_list[[12]]$start_station_id = as.numeric(df_list[[12]]$start_station_id)
df_list[[12]]$end_station_id = as.numeric(df_list[[12]]$end_station_id)

# Bind data frames by rows
df = bind_rows(df_list)

# Glimpse at df
glimpse(df)
```
### Format Data Structure

```{r format-data-struct}
df = df %>%
  # Extract day and month from start_time
  mutate(start_month = sub(" .*", "", start_time)) %>%
  mutate(start_month = as.POSIXct(start_month)) %>%
  mutate(start_day = weekdays(start_month)) %>%
  mutate(start_month = format(start_month, "%B")) %>%

  # Extract hour from start_time
  mutate(start_hour = sub("^.{11}", "", start_time)) %>%
  mutate(start_hour = str_extract(start_hour, "^[0-9]{2}")) %>%

  # Convert duration_sec to duration_min
  mutate(duration_min = duration_sec / 60) %>%

  # Convert characters to factors
  mutate_if(is.character, as.factor) %>%

  # Remove columns not used in the analysis
  select(-c(start_time, end_time, start_station_id, end_station_id, bike_id, duration_sec))

# Glimpse at cleaned df
glimpse(df)
```

*** 

## **Data Analysis**
### Bike Usage by Month and by Day in 2018

```{r usage-by-month-and-day}
# Organize month
df$start_month = factor(df$start_month,
                        levels = c("January", "February", "March", "April", "May", "June",
                                   "July", "August", "September", "October", "November", "December"))

# Count usage by month
month_counts = table(df$start_month)

# Plot bar graph
bar = barplot(month_counts,
              ylim = c(0, 220000),
              xlab = "Months in 2018",
              ylab = "Number of Rides",
              main = "Bike Usage by Month",
              cex.names = 0.8,
              cex.axis = 0.8)
text(x = bar, y = month_counts, # Add labels
     label = month_counts, pos = 3, cex = 0.8, col = "red")

# Organize day of the week
df$start_day = factor(df$start_day,
                      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

# Count usage by day
day_counts = table(df$start_day)

# Plot bar graph
bar = barplot(day_counts,
              ylim = c(0, 350000),
              xlab = "Day of the Week",
              ylab = "Number of Rides",
              main = "Bike Usage by Day of the Week",
              cex.axis = 0.8,
              xaxt = "n")
text(x = bar, y = day_counts, # Add labels
     label = day_counts, pos = 3, cex = 0.8, col = "red")
text(x = bar, y = -6000, cex = 0.8, # Rotate x-axis labels
     labels = names(day_counts), srt = 45, adj = 1, xpd = TRUE)
```

### Bike Usage by Customers and Subscribers

```{r usage-by-cust-subs}
# Plot bar graph
ggplot(df, aes(start_month)) +
  geom_bar() +
  facet_grid(. ~ user_type) +
  xlab("Months in 2018") +
  ylab("Number of Rides") +
  ggtitle("Comparison of Usage Between Customers and Subscribers") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis 
```

### Distribution of Ride Duration

```{r dist-ride-duration}
# Plot histogram
ggplot(df, aes(duration_min)) +
  geom_histogram(binwidth = 2, alpha = 0.5) +
  scale_x_continuous(limits = c(0, 100),
                     breaks = seq(0, 100, by = 2)) +
  xlab("Ride Duration in Minutes") +
  ylab("Number of Rides") +
  ggtitle("Duration of Rides") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) # Rotate x-axis labels
```

### Bike Usage by Hour

```{r usage-by-hour}
# Organize usage by hour
hour = df %>%
  group_by(start_hour) %>%
  summarise(total = n())
```










